---
title: "w241: Experiments and Causality"
subtitle: "Unit 3"
author: "David Reiley, David Broockman, D. Alex Hughes"
institute: "UC Berkeley, School of Information"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(data.table)

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 300)


theme_set(theme_minimal())
```

class: inverse, center, middle

# Sampling Distribution and Randomization Inference

---

# Standard Errors 

- Standard deviation of the sampling distribution 
- How spread out is the sampling distribution? 
- How large are the typical chance differences? 
- Later, we'll examine statistical power 
  - The spread of the sampling distribution is the standard error 
  - In what kinds of experiments are large and small differences likely to arise by chance? 
  
---

# Sampling Distributions and RI

- Groups may differ by chance, even if the treatment has no effect.
  - How much would the groups differ if the treatment had no effect?
  - How large of an "effect estimate" would we reach by chance?
- Distribution of estimates one would reach if treatment had no effect.
  - How likely is this estimate to have just arisen by
chance?
- Similar to observational studies, but:
  - Intuition easy to see in experiments.
  - Testing a hypothesis about our sample, not a population.
  - Example code to walk through intuition on slides to follow  
  
---

# Example: An Experiment with no Effect

- Does eating soybeans affect estrogen levels?
- 40 individuals: 20 men, 20 women.
- Simulate the potential outcomes of the control group. 
- Simulate the potential outcomes of the treatment group. 
- A simulated experiment with no effect.

---

```{r make potential outcomes} 
group <- c(rep("Man",20),rep("Woman",20))

po_control <- c(
  seq(from = 1, to = 20), 
  seq(from = 51, to = 70)
  )

## Suppose there is no effect. 
## Then, the potential outcomes to control are equal 
## to the potential outcomes to treatment. 

po_treatment <- po_control + 0

d <- data.frame(
  'Control'   = po_control,
  'Treatment' = po_treatment, 
  'group'        = group
  )
```

---

```{r print head of dataframe}
d %>% 
  head()
```


---

```{r plot outcomes, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
d %>% 
  pivot_longer(cols = c('Control', 'Treatment')) %>% 
  ggplot() + 
  aes(x = value, color = group, linetype = name) + 
  geom_density() + 
  facet_wrap(facets = vars(name)) + 
  labs(
    title = "Distribution of Potential Outcomes", 
    subtitle = "Man identifiers have lower estrogen",
    x = 'Potential Outcomes', 
    y = 'Density', 
    color = 'Identity Group', 
    linetype = 'Potential Outcome'
  )
```

---

# Random Assignment 

- Define function to randomly assign units to treatment and control.
- Randomly pick 20 for treatment and 20 for control. 
- Concatenate the two vectors.
- Get a different vector when you run it again.

```{r define randomization function}
randomize <- function(units_per_group) { 
  ## an (unnecessary) function to randomize units into 
  ## treatment and control 
  ## ---
  ## args: 
  ##  - units_per_group: how many zero and one should be returned
  
  assignment_vector <- rep(c('Control', 'Treatment'), each = units_per_group)
  sample(assignment_vector)
} 
```

---

# Random Assignment

```{r demonstrate randomizaton function}

randomize(units_per_group = 4) 
randomize(units_per_group = 4) 
```

---

# Realized Outcomes 

- Treatment outcome for those randomized to treatment and control outcome for those randomized to control.
- Assign for each person in the vector.
- Same because we had an experiment with no effect.
- R code is often written in a compact manner; could also have been done separately for each group.
- Why are we doing this when there is no treatment effect?
- Because it should also work when there is one. We're looking at what happens when we randomly assign people to control and treatment groups.

---

# Realized Outcomes

```{r assign treatment}
treatment_assigned <- randomize() 

outcomes <- po_treatment * I(treatment_assigned == "Treatment") + 
  po_control * I(treatment_assigned == "Control")

outcomes
```

---

# Function to Estimate the Average Treatment Effect (ATE) 

- Subtract the mean outcome for the control group from the mean outcome of the treatment group.
- How much higher is the average in the treatment group versus the control group?
- We may have randomly selected someone with a higher or lower level of estrogen.
- Even though we know the effect is 0, we see chance differences.

---

# Function to Estimate the Average Treatment Effect (ATE) 

```{r define estimate ate function}
estimate_ate <- function(y_values, treatment) { 
  
  treatment_group_mean <- mean(y_values[treatment == 'Treatment'])
  control_group_mean   <- mean(y_values[treatment == 'Control'])
  
  ate <- treatment_group_mean - control_group_mean
  
  return(
    list(
      "tg_mean" = treatment_group_mean, 
      "cg_mean" = control_group_mean, 
      "ate" = ate)
  )
}
```

---

```{r, run estimate ate function once}
## In fact, there is _no_ effect, but... sampling! 
estimate_ate(y_values = outcomes, treatment = treatment_assigned) 

## To pull a single part of this, because it is a list, R indexes with .[[
estimate_ate(y_values = outcomes, treatment = treatment_assigned)[['ate']]
```

---

class: inverse, center, middle

# The Null Hypothesis

---

# Rhetorical Posture of the Null Hypothesis
- You want to argue against a skeptic that a treatment has an effect.
- Assume the skeptic is right.

> Treatment has no effect.

- What is the chance that we would see this estimate by chance in that scenario?

> This is p-value.

- We’ll see where it comes from visually.

---

# Average Size of the Difference

- Simulate this a few times to get a sense of how much our treatment effect estimate would vary by chance.
- We created an estimate function with the outcomes and the treatment group.
- Outcome vector will look the same regardless of the treatment vector.

```{r a few experiments}
treatment_assigned_one <- randomize(units_per_group = 20)
estimate_ate(y_values = outcomes, treatment = treatment_assigned_one)[['ate']]

treatment_assigned_two <- randomize(units_per_group = 20)
estimate_ate(y_values = outcomes, treatment = treatment_assigned_two)[['ate']]

treatment_assigned_three <- randomize(units_per_group = 20)
estimate_ate(y_values = outcomes, treatment = treatment_assigned_three)[['ate']]
```

---

# Outcome With Different Assignments

- Similar to re-sampling from a population.
- Re-randomizing from within the original population. Testing the null hypothesis from within the sample we already have.
- Re-shuffle the 40 people between treatment and control. Assuming the treatment effect for everyone is zero. 
- **Sharp null hypothesis**: For every unit, there is no effect.
- Repeat this process to generate a synthetic distribution of effects if the sharp null hypothesis *were true*. 

- Randomly sample the vector of assignments 5,000 times to generate an unbiased sample of all the effects.
- Literally, replicate 5,000 times, and save to a vector.

---

```{r}
## going to move the randomization inside the `estimate_ate` function 
## for compactness 

sharp_null <- replicate(
  n = 5000, 
  expr = estimate_ate(
    y_value = outcomes, 
    treatment = randomize(units_per_group = 20))[['ate']]
)
```

---

```{r sharp null distribution, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
ggplot() + 
  aes(x = sharp_null) + 
  geom_histogram() + 
  labs(
    title = "Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```

---

# Size of the Observed Difference

- The p-value.
- How often did I get a randomization under the sharp null where the estimate was larger than my actual estimate?
- For each, is it larger than the average treatment effect estimate?
- This is a sampling distribution.
- How big is my estimate relative to the distribution of estimates?

---

# Size of the Observed Difference 

In this particular case, 

```{r estimate p-value}
experimental_randomization <- randomize(units_per_group = 20)
experimental_ate <- estimate_ate(
  y_values = outcomes, 
  treatment = experimental_randomization)[['ate']]

sharp_null <- replicate(
  n = 5000, 
  expr = estimate_ate(
    y_value = outcomes, 
    treatment = randomize(units_per_group = 20))[['ate']]
)

mean(abs(sharp_null) > abs(experimental_ate))
```

---

# Size of the Observed Difference 

```{r make no effect histogram}
histogram_no_effect <- ggplot() + 
  aes(x = sharp_null) + 
  geom_histogram() + 
  geom_vline(xintercept = experimental_ate, color = 'darkorange') + 
  labs(
    title = "No Effect: Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```

---

# Size of the Observed Difference 

```{r plot sharpnull and experimental ate, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
histogram_no_effect
```

---

class: inverse, center, middle

# P-Values and Hypothesis Tests

---

# P-Values 

- If the treatment had no effect, how likely is it that the data would generate  a difference this extreme, *just by chance*?
- What is the difference between the mean in the control and treatment groups?
- Different from how likely it is the treatment has an effect
- Convention is to reject the null with p-value under 0.05.
- p-values don't tell you for sure that the treatment has an effect.
- They just tell you how likely it is you would have gotten that result by chance.
- The sampling distribution tells us how large the differences are we find by chance.
- Can find p-values < 0.05 even when the null hypothesis is correct.

---

# Simulating an Experiment with a Large Effect

- Vector of outcomes and control
- 40-row table with potential outcomes in control and treatment. 
- This time, with a difference of 25

---

# Simulating an Experiment with a Large Effect

```{r big effect}
po_control   <- c(1:20, 51:70)
po_treatment <- po_control + 25

treatment_assigned <- randomize(units_per_group = 20)

outcomes <- po_treatment * I(treatment_assigned == "Treatment") + 
  po_control * I(treatment_assigned == "Control")

experimental_ate_big_effect <- estimate_ate(
  y_values = outcomes, 
  treatment = treatment_assigned
  )[['ate']]
```

---

# Simulating an Experiment with a Large Effect

```{r}
sharp_null_big_effect <- replicate(
  n = 5000, 
  expr = estimate_ate(
    y_values = outcomes, 
    treatment = randomize(units_per_group = 20))[['ate']]
  )
```

---

# Simulating an Experiment with a Large Effect

```{r make big effect histogram} 
histogram_big_effect <- ggplot() + 
  aes(x = sharp_null_big_effect) + 
  geom_histogram() + 
  geom_vline(xintercept = experimental_ate_big_effect, color = 'darkorange') + 
  labs(
    title = "Big Effect: Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```

---

# Simulating an Experiment with a Large Effect

```{r plot sharpnull and experimental ate big effect experiment, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
histogram_big_effect
```

---

# Simulating an Experiment with a Large Effect

```{r}
mean(abs(sharp_null_big_effect) > abs(experimental_ate_big_effect))
```

---

# Compare Big Effect and No Effect Sharp Null Disributions 

```{r, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'}
histogram_big_effect / histogram_no_effect
```

---

class: inverse, center, middle

# Statistical Power 

---

# Detecting Non-Zero Treatmetn Effects

Suppose the treatment effect is 10. 

---

# Create Whole Study Function 

```{r define simulate study function} 
simulate_study <- function(effect_size) { 
  # requires all functions to date
  
  # generate world
  po_control   <- c(1:20, 51:70)
  po_treatment <- po_control + effect_size
  
  # assign treatment and measure outcomes
  treatment_assigned <- randomize(20)
  outcomes <- po_treatment * I(treatment_assigned == "Treatment") + 
    po_control * I(treatment_assigned == "Control") 
  
  # estimate ate   
  estimated_ate <- estimate_ate(y_values = outcomes, treatment = treatment_assigned)[['ate']]
  
  # generate sharp null distribution
  sharp_null <- replicate(
    n = 100, 
    expr = estimate_ate(y_values = outcomes, treatment = randomize(20)
    )[['ate']]
  )
  
  p_value <- mean(abs(sharp_null) > abs(estimated_ate))
  return(list(
    'estimated_ate' = estimated_ate,
    'mean_sharp_null' = mean(sharp_null), 
    'p_value' = p_value)
  )
}

simulate_study(effect_size = 5)
```

---

# Generate P-Values, Effect Size: 0 

```{r conduct power analysis effect size 0} 
## notice: we now have two loops: 
  ## - We're running 1,000 simulations;
  ## - In each simulation, there are 1,000 sharp nulls drawn out
  ## - So get some coffee if you're running this at home

distribution_of_p_values_0 <- replicate(
  n = 200, 
  expr = simulate_study(effect_size = 0)[['p_value']]
)
```

---

# Plot Distribution of P-Values, Effect Size: 0

```{r}
histogram_pvalues_0 <- ggplot() + 
  aes(distribution_of_p_values_0) + 
  geom_histogram() + 
  labs(
    title = "Effect Size: 0", 
    subtitle = paste0("Percent P-Values < 0.05: ", round(mean(distribution_of_p_values_0 < 0.05) * 100), "%"), 
    x = "P-Value")
```

# Generate P-Value, Effect Size 10 

```{r conduct power analysis effect size 10}
distribution_of_p_values_10 <- replicate(
  n = 200, 
  expr = simulate_study(effect_size = 10)[['p_value']]
)
```

---

# Plot Distribution of P-Values Effect Size: 10 

```{r plot pvalues for effect size 10}
histogram_pvalues_10 <- ggplot() + 
  aes(x = distribution_of_p_values_10) + 
  geom_histogram() + 
  labs(
    title = "Effect Size: 10", 
    subtitle = paste0("Percent P-Values < 0.05: ", round(mean(distribution_of_p_values_10 < 0.05) * 100), "%"), 
    x = 'P-Value'
  )
```

---

# Generate P-Value, Effect Size 20

```{r conduct power analysis effect size 20}
distribution_of_p_values_20 <- replicate(
  n = 200, 
  expr = simulate_study(effect_size = 20)[['p_value']]
)
```

```{r plot pvalues for effect size 20}
histogram_pvalues_20 <- ggplot() + 
  aes(x = distribution_of_p_values_20) + 
  geom_histogram() + 
  labs(
    title = "Effect Size: 20",
    subtitle = paste0("Percent P-Values < 0.05: ", round(mean(distribution_of_p_values_20 < 0.05) * 100), "%"), 
    x = 'P-Value')
    
```

# Plot Distributions of P-Values

```{r plot all power histograms, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'}
histogram_pvalues_0 / histogram_pvalues_10 / histogram_pvalues_20
```