---
title: "w241: Experiments and Causality"
subtitle: "Unit 3"
author: "David Reiley, David Broockman, D. Alex Hughes"
institute: "UC Berkeley, School of Information"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(data.table)

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 300)


theme_set(theme_minimal())
```

class: inverse, center, middle

# Sampling Distribution and Randomization Inference

---

# Standard Errors 

- Standard deviation of the sampling distribution 
- How spread out is the sampling distribution? 
- How large are the typical chance differences? 
- Later, we'll examine statistical power 
  - The spread of the sampling distribution is the standard error 
  - In what kinds of experiments are large and small differences likely to arise by chance? 
  
---

# Sampling Distributions and RI

- Groups may differ by chance, even if the treatment has no effect.
  - How much would the groups differ if the treatment had no effect?
  - How large of an "effect estimate" would we reach by chance?
- Distribution of estimates one would reach if treatment had no effect.
  - How likely is this estimate to have just arisen by
chance?
- Similar to observational studies, but:
  - Intuition easy to see in experiments.
  - Testing a hypothesis about our sample, not a population.
  - Example code to walk through intuition on slides to follow  
  
---

# Example: An Experiment with no Effect

- Does eating soybeans affect estrogen levels?
- 40 individuals: 20 men, 20 women.
- Simulate the potential outcomes of the control group. 
- Simulate the potential outcomes of the treatment group. 
- A simulated experiment with no effect.

---

```{r make potential outcomes} 
group <- c(rep("Man",20),rep("Woman",20))

po_control <- c(
  seq(from = 1, to = 20), 
  seq(from = 51, to = 70)
  )

## Suppose there is no effect. 
## Then, the potential outcomes to control are equal 
## to the potential outcomes to treatment. 

po_treatment <- po_control + 0

d <- data.frame(
  'Control'   = po_control,
  'Treatment' = po_treatment, 
  'group'        = group
  )
```

---

```{r print head of dataframe}
d %>% 
  head()
```


---

```{r plot outcomes, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
d %>% 
  pivot_longer(cols = c('Control', 'Treatment')) %>% 
  ggplot() + 
  aes(x = value, color = group, linetype = name) + 
  geom_density() + 
  facet_wrap(facets = vars(name)) + 
  labs(
    title = "Distribution of Potential Outcomes", 
    subtitle = "Man identifiers have lower estrogen",
    x = 'Potential Outcomes', 
    y = 'Density', 
    color = 'Identity Group', 
    linetype = 'Potential Outcome'
  )
```

---

# Random Assignment 

- Define function to randomly assign units to treatment and control.
- Randomly pick 20 for treatment and 20 for control. 
- Concatenate the two vectors.
- Get a different vector when you run it again.

```{r define randomization function}
randomize <- function(units_per_group) { 
  ## an (unnecessary) function to randomize units into 
  ## treatment and control 
  ## ---
  ## args: 
  ##  - units_per_group: how many zero and one should be returned
  
  assignment_vector <- rep(c('Control', 'Treatment'), each = units_per_group)
  sample(assignment_vector)
} 
```

---

# Random Assignment

```{r demonstrate randomizaton function}

randomize(units_per_group = 4) 
randomize(units_per_group = 4) 
```

---

# Realized Outcomes 

- Treatment outcome for those randomized to treatment and control outcome for those randomized to control.
- Assign for each person in the vector.
- Same because we had an experiment with no effect.
- R code is often written in a compact manner; could also have been done separately for each group.
- Why are we doing this when there is no treatment effect?
- Because it should also work when there is one. We're looking at what happens when we randomly assign people to control and treatment groups.

---

# Realized Outcomes

```{r assign treatment}
treatment_assigned <- randomize() 

outcomes <- po_treatment * I(treatment_assigned == "Treatment") + 
  po_control * I(treatment_assigned == "Control")

outcomes
```

---

# Function to Estimate the Average Treatment Effect (ATE) 

- Subtract the mean outcome for the control group from the mean outcome of the treatment group.
- How much higher is the average in the treatment group versus the control group?
- We may have randomly selected someone with a higher or lower level of estrogen.
- Even though we know the effect is 0, we see chance differences.

---

# Function to Estimate the Average Treatment Effect (ATE) 

```{r define estimate ate function}
estimate_ate <- function(y_values, treatment) { 
  
  treatment_group_mean <- mean(y_values[treatment == 'Treatment'])
  control_group_mean   <- mean(y_values[treatment == 'Control'])
  
  ate <- treatment_group_mean - control_group_mean
  
  return(
    list(
      "tg_mean" = treatment_group_mean, 
      "cg_mean" = control_group_mean, 
      "ate" = ate)
  )
}
```

---

```{r, run estimate ate function once}
## In fact, there is _no_ effect, but... sampling! 
estimate_ate(y_values = outcomes, treatment = treatment_assigned) 

## To pull a single part of this, because it is a list, R indexes with .[[
estimate_ate(y_values = outcomes, treatment = treatment_assigned)[['ate']]
```

---

class: inverse, center, middle

# The Null Hypothesis

---

# Rhetorical Posture of the Null Hypothesis
- You want to argue against a skeptic that a treatment has an effect.
- Assume the skeptic is right.

> Treatment has no effect.

- What is the chance that we would see this estimate by chance in that scenario?

> This is p-value.

- We’ll see where it comes from visually.

---
# Average Size of the Difference 

### Because this demonstration is based on a stochastic simulation, the *specific* values that are in the slides might not match what we're narrating aloud. 

### What we're reading aloud are the results for the trial that we conducted. 

---

# Average Size of the Difference

- Simulate this a few times to get a sense of how much our treatment effect estimate would vary by chance.
- We created an estimate function with the outcomes and the treatment group.
- Outcome vector will look the same regardless of the treatment vector.

```{r a few experiments}
treatment_assigned_one <- randomize(units_per_group = 20)
estimate_ate(y_values = outcomes, treatment = treatment_assigned_one)[['ate']]

treatment_assigned_two <- randomize(units_per_group = 20)
estimate_ate(y_values = outcomes, treatment = treatment_assigned_two)[['ate']]

treatment_assigned_three <- randomize(units_per_group = 20)
estimate_ate(y_values = outcomes, treatment = treatment_assigned_three)[['ate']]
```

---

# Outcome With Different Assignments

- Similar to re-sampling from a population.
- Re-randomizing from within the original population. Testing the null hypothesis from within the sample we already have.
- Re-shuffle the 40 people between treatment and control. Assuming the treatment effect for everyone is zero. 
## Sharp null hypothesis: For every unit, there is no effect.

- Repeat this process to generate a synthetic distribution of effects if the sharp null hypothesis *were true*. 

- Randomly sample the vector of assignments 5,000 times to generate an unbiased sample of all the effects.
- Literally, replicate 5,000 times, and save to a vector.

---

```{r}
## going to move the randomization inside the `estimate_ate` function 
## for compactness 

sharp_null <- replicate(
  n = 5000, 
  expr = estimate_ate(
    y_value = outcomes, 
    treatment = randomize(units_per_group = 20))[['ate']]
)
```

```{r, echo = FALSE}
experimental_ate <- 1.3
```

---

```{r sharp null distribution, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
ggplot() + 
  aes(x = sharp_null) + 
  geom_histogram() + 
  labs(
    title = "Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```

---

```{r sharp null distribution with abline, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
ggplot() + 
  aes(x = sharp_null) + 
  geom_histogram() + 
  geom_vline(xintercept = experimental_ate, color = 'darkorange') + 
  labs(
    title = "Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```


---

# Size of the Observed Difference

- The p-value.
- How often did I get a randomization under the sharp null where the estimate was larger than my actual estimate?
- For each, is it larger than the average treatment effect estimate?
- This is a sampling distribution.
- How big is my estimate relative to the distribution of estimates?

---

# Size of the Observed Difference 

In this particular case, 

```{r estimate p-value}
experimental_randomization <- randomize(units_per_group = 20)

sharp_null <- replicate(
  n = 5000, 
  expr = estimate_ate(
    y_value = outcomes, 
    treatment = randomize(units_per_group = 20))[['ate']]
)

mean(abs(sharp_null) > abs(experimental_ate))
```

---

# Size of the Observed Difference 

```{r make no effect histogram}
histogram_no_effect <- ggplot() + 
  aes(x = sharp_null) + 
  geom_histogram() + 
  geom_vline(xintercept = experimental_ate, color = 'darkorange') + 
  labs(
    title = "No Effect: Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```

---

# Size of the Observed Difference 

```{r plot sharpnull and experimental ate, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
histogram_no_effect
```

---

class: inverse, center, middle

# P-Values and Hypothesis Tests

---

# P-Values 

- If the treatment had no effect, how likely is it that the data would generate  a difference this extreme, *just by chance*?
- What is the difference between the mean in the control and treatment groups?
- Different from how likely it is the treatment has an effect
- Convention is to reject the null with p-value under 0.05.
- p-values don't tell you for sure that the treatment has an effect.
- They just tell you how likely it is you would have gotten that result by chance.
- The sampling distribution tells us how large the differences are we find by chance.
- Can find p-values < 0.05 even when the null hypothesis is correct.

---

# Simulating an Experiment with a Large Effect

- Vector of outcomes and control
- 40-row table with potential outcomes in control and treatment. 
- This time, with a difference of 25

---

# Simulating an Experiment with a Large Effect

```{r big effect}
po_control   <- c(1:20, 51:70)
po_treatment <- po_control + 25

treatment_assigned <- randomize(units_per_group = 20)

outcomes <- po_treatment * I(treatment_assigned == "Treatment") + 
  po_control * I(treatment_assigned == "Control")
outcomes

experimental_ate_big_effect <- estimate_ate(
  y_values = outcomes, 
  treatment = treatment_assigned
  )[['ate']]
experimental_ate_big_effect
```

---

# Simulating an Experiment with a Large Effect

```{r}
sharp_null_big_effect <- replicate(
  n = 5000, 
  expr = estimate_ate(
    y_values = outcomes, 
    treatment = randomize(units_per_group = 20))[['ate']]
  )
```

---

# Simulating an Experiment with a Large Effect

```{r make big effect histogram} 
histogram_big_effect <- ggplot() + 
  aes(x = sharp_null_big_effect) + 
  geom_histogram() + 
  geom_vline(xintercept = experimental_ate_big_effect, color = 'darkorange') + 
  labs(
    title = "Big Effect: Distribution of Treatment Effects Under Sharp Null", 
    subtitle = "Distribution is Centered at Zero, And Symmetric", 
    x = "Estimated Average Treatment Effect", 
    y = "Count"
  )
```

---

# Simulating an Experiment with a Large Effect

```{r plot sharpnull and experimental ate big effect experiment, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'} 
histogram_big_effect
```

---

# Simulating an Experiment with a Large Effect

```{r}
mean(abs(sharp_null_big_effect) > abs(experimental_ate_big_effect))
```

---

# Compare Big Effect and No Effect Sharp Null Disributions 

```{r, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'}
histogram_big_effect / histogram_no_effect
```

---

class: inverse, center, middle

# Statistical Power 

---

# Detecting Non-Zero Treatment Effects

Suppose the treatment effect is 10. 

---

# Create Whole Study Function 

```{r define simulate study function} 
simulate_study <- function(effect_size) { 
  # generate world
  po_control   <- c(1:20, 51:70)
  po_treatment <- po_control + effect_size
  
  # assign treatment and measure outcomes
  treatment_assigned <- randomize(20)
  outcomes <- po_treatment * I(treatment_assigned == "Treatment") + 
    po_control * I(treatment_assigned == "Control") 
  
  # estimate ate   
  estimated_ate <- estimate_ate(y_values = outcomes, treatment = treatment_assigned)[['ate']]
  
  # generate sharp null distribution
  sharp_null <- replicate(
    n = 100, 
    expr = estimate_ate(y_values = outcomes, treatment = randomize(20))[['ate']])
  
  p_value <- mean(abs(sharp_null) > abs(estimated_ate))
  return(list(
    'estimated_ate' = estimated_ate,
    'mean_sharp_null' = mean(sharp_null), 
    'p_value' = p_value)
  )
}
```

---

# Simulate Study, Effect Size: 10

```{r conduct power analysis effect size 10}
## notice: we now have two loops: 
  ## - We're running 500 simulations;
  ## - In each simulation, there are 1,000 sharp nulls drawn out
  ## - So get some coffee if you're running this at home

distribution_of_p_values_10 <- replicate(
  n = 500, 
  expr = simulate_study(effect_size = 10)[['p_value']]
)
```

---

# Power for 10 Unit Effect

```{r plot distribution for effect size of 10, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'}
ggplot() + 
  aes(x = distribution_of_p_values_10) + 
  geom_density(color = '#00A598') + 
  geom_vline(xintercept = 0.05, color = 'darkorange') + 
  labs(
    title = 'Distribution of P-Values: Effect Size = 10', 
    subtitle = 'Many tests generate p-values larger than 0.05', 
    x = 'P-Values', 
    y = 'Density'
  )
```

---

# Power Sim: 0 & 20 Unit Effect

```{r compute power analysis for zero effect and large effect } 
distribution_of_p_values_0 <- replicate(
  n = 500, 
  expr = simulate_study(effect_size = 0)[['p_value']]
)

distribution_of_p_values_20 <- replicate(
  n = 500, 
  expr = simulate_study(effect_size = 20)[['p_value']]
)
```

---

# Power Curves for All Effects

```{r plot distributions, message = FALSE, echo = FALSE, dev='svglite', fig.height=4, fig.align='center'}
d <- data.frame(
  p_values = c(distribution_of_p_values_0, distribution_of_p_values_10, distribution_of_p_values_20), 
  group = c(rep('No Effect', length(distribution_of_p_values_0)), 
            rep('10 Unit Effect', length(distribution_of_p_values_10)), 
            rep('20 Unit Effect', length(distribution_of_p_values_20)))
)

d %>% 
  ggplot() + 
  aes(x = p_values, color = group) + 
  geom_density() + 
  scale_color_manual(values = c('#00A598', '#003262', '#FDB515')) + 
  geom_vline(xintercept = 0.05, color = 'darkorange') + 
  labs(
    title = 'Distribution of P-Values for Different Effect Sizes', 
    subtitle = 'Left Shift Rejects Null More Frequently', 
    color = 'Effect Size',
    x = 'P-values', 
    y = 'Density'
  )
```

---

# Increasing Statistical Power

## Power Increases With: 

- Size of the effect -- *larger effects are easier to detect!*
- Square root of the sample size, $\sqrt{N}$. 
  - To detect an effect twice as small (or equivalently half as large) requires a sample size 4 times larger; 
- Precision of the measurement 
- Reduction of variance within groups (e.g. removing individuals pre-test; or block randomizing)

## Statistical Power:

> "The probability that a particular {experiment design & measurment & test} will reject the null hypothesis in a world where it *should* reject that null hypothesis."

---

# Concentrated Tests 

## Suppose the FDA is testing the effect of soybeans on estrogen

-  **Study One**: Give one soybean to 1,000,000 people. 
- **Study Two**: Give 10 soybeans to 10,000 people. 
  - If there is a linear effect of soybeans, then these two design have equivalent power
  - However, *Study Two* has used 1/10 as many soybeans in the study.=
    - If the input is the expensive part of the experiment, then this saves cost on the input 
    - If the recruitment of subjects is the expensive part of the experiment, then this has *also* saved cost on the recruitment. 
- (**Study Three**): Give 100 soybeans to 100 people has the same power as the above two experiments as well!      

---

# Concentrated Tests 

- Often, it is a good idea to decrease the sample size and give a higher "dosage" to the treatment group
- Concentrated tests increase statistical power by exposing a smaller number of people to a larger dose of treatment. 

---

# Decreasing Statistical Power 

## Power Decreasese With: 

- Larger amounts of variation in the measured outcomes
  - More diverse populations create more differences in baseline differences; relative to the effect size, this "mutes" the ability to measure an effect 
  - More "noise" in the measurement raises the "floor" of what one must detect to look different from that noise; precise measurements are preferred to imprecise measurements
- Standard deviation, $\sigma$, of the outcome

## Key Concept: 

- The ratio of the true treatment effect to the standard error of the estimated effect: 

$$
  \text{test statistic} = \frac{\hat{\tau}}{SE(\hat{\tau})} =  \frac{\hat{\tau}}{\left(\frac{\sigma_{\hat{\tau}}}{\sqrt{N}}\right)}
$$

---

class: inverse, center, middle
# Recap of the Week 

---

# Recap of the Week, Part I 

## Sampling Distribution 

- A **sharp null sampling distribution** is a distribution of estimates that we would receive by chance if there really were *no effect*
  - The "sharp null distribution" then simulates the range of outcomes our experiment and estimate system can produce, even when there is *no effect* 
  - The proportion of these simulations that are *more-extreme* than the treatment effect observed in the experiment is the **randomization inference p-value**. 
  
---

# Recap of the Week, Part II
  
## P-Values 

- **P-values** provide a statement about $P(\text{data} | \text{sharp-null is true})$. 
- What we would *ideally* like to know is $P(\text{alternative is true} | \text{data})$, but this isn't provided by randomization inference, or Frequentest inference. 

---

# Recap of the Week, Part III

## Statistical Power 

- Simply increasing sample size of an experiment can improve **statistical power** which is the *the probability that a test will reject the sharp-null hypothesis when the sharp null is* actually *true*. 
- Careful design can also improve statistical power. 