---
title: "w241: Experiments and Causality"
subtitle: "Unit 4"
author: "David Reiley, David Broockman, D. Alex Hughes"
institute: "UC Berkeley, School of Information"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(data.table)

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 300)


theme_set(theme_minimal())
```

# Blocking

- Hard to know if it was due to chance when there are large differences between treatment and control.
- Need to reduce the size of the differences that can arise by chance.
- Increase statistical power given an experiment with same sample and effect size.
- If some variables are related to the outcome, restrict ourselves to randomizations that keep treatment and control similar.

---

# Men and Women Group Variable

```{r make potential outcomes} 
group <- c(rep("Man",20),rep("Woman",20))

po_control <- c(
  seq(from = 1, to = 20), 
  seq(from = 51, to = 70)
  )

## Suppose there is no effect. 
## Then, the potential outcomes to control are equal 
## to the potential outcomes to treatment. 

po_treatment <- po_control + 0

d <- data.table(
  'Control'   = po_control,
  'Treatment' = po_treatment, 
  'group'        = group
  )
```

```{r}
d %>% head()
d %>% tail() 
```

--- 

# Randomization 

```{r define randomization function}
randomize <- function(units_per_group) { 
  ## an (unnecessary) function to randomize units into 
  ## treatment and control 
  ## ---
  ## args: 
  ##  - units_per_group: how many zero and one should be returned
  
  assignment_vector <- rep(c('Control', 'Treatment'), each = units_per_group)
  sample(assignment_vector)
} 
```

```{r}
table(
  'Sex' = d$group, 
  'Assignment' = randomize(20)
)

table(
  'Sex' = d$group, 
  'Assignment' = randomize(20)
)

table(
  'Sex' = d$group, 
  'Assignment' = randomize(20)
)
```

```{r} 

block_randomize <- function(size) { 
  ## this function will be executed /within/ the  data.table that 
  ## holds the data. It could be run outside, but the assignment 
  ## in place that data.table provides make it clean inside.
  conditions <- c('Control', 'Treatment')
  
  if(size %% 2 == 0) { 
    ## if there are an even number of units in each block this is easy
    urn <- rep(conditions, times = size/2)  
  } else if(size %% 2 == 1) { 
    ## if there are an odd number, then produce conditions to the 
    ## nearest even number that is less than the number of units 
    ## then add one more assignment condition, sampled at random
    urn <- c(rep(conditions, times = (size/2) - 0.5), sample(conditions, size = 1))
  } 
  
  ## now, shuffle it up and (implicitly) return the shuffled sequence
  sample(urn)
}
  
  
```

```{r run randomize}
d[ , block_assignment := block_randomize(size = .N), by = group]

d[ , table(
  'Condition' = block_assignment, 
  'Sex' = group
  )]
```

# Men and Women Group Variable (contd)
>randomize.blocked <- function(){
+ c(
+ sample(c(rep(0,10),rep(1,10))), #group A + sample(c(rep(0,10),rep(1,10))) #group B +)
+}
>#now groups are always balanced
>table(group, randomize.blocked())
group 0 1
Man 10 10
Woman 10 10
>table(group, randomize.blocked())
group 0 1
Man 10 10
Woman 10 10
# Simulate the Normal Study
>sim.normal.study <- function(){
+ po.control <- c(seq(from = 1, to = 20), seq(from = 51, to = 70)) + po.treatment <- po.control
+ treatment <- randomize()
+ outcomes <- po.treatment * treatment + po.control * (1–treatment) + ate <- est.ate(outcomes, treatment)
+ n.women.treatment <- table(group, treatment)[2,2]
+ return(list(ate = ate, n.women.treatment = n.women.treatment)) +}
>results <- t(replicate(1000, sim.normal.study()))
>plot(results)
 https://learn.datascience.berkeley.edu/oyster/player/latest/index.html?segmentUuid=7c4d54ad-a6b1-46bb-a1cb-e24a9815bf3a&print=true Page 2 of 3

Blocking 5/19/21, 4:52 PM
 https://learn.datascience.berkeley.edu/oyster/player/latest/index.html?segmentUuid=7c4d54ad-a6b1-46bb-a1cb-e24a9815bf3a&print=true Page 3 of 3
