---
title: "w241: Experiments and Causality"
subtitle: "Unit 4"
author: "David Reiley, David Broockman, D. Alex Hughes"
institute: "UC Berkeley, School of Information"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(data.table)

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 300)


theme_set(theme_minimal())
```

# Blocking

- Hard to know if it was due to chance when there are large differences between treatment and control.
- Need to reduce the size of the differences that can arise by chance.
- Increase statistical power given an experiment with same sample and effect size.
- If some variables are related to the outcome, restrict ourselves to randomizations that keep treatment and control similar.

---

# Men and Women Group Variable

```{r make potential outcomes} 
group <- c(rep("Man",20),rep("Woman",20))

po_control <- c(
  seq(from = 1, to = 20), 
  seq(from = 51, to = 70)
  )

## Suppose there is no effect. 
## Then, the potential outcomes to control are equal 
## to the potential outcomes to treatment. 

po_treatment <- po_control + 0

d <- data.table(
  'Control'   = po_control,
  'Treatment' = po_treatment, 
  'group'        = group
  )
```

```{r}
d %>% head()
d %>% tail() 
```

--- 

# Randomization 

```{r define randomization function}
randomize <- function(units_per_group) { 
  ## an (unnecessary) function to randomize units into 
  ## treatment and control 
  ## ---
  ## args: 
  ##  - units_per_group: how many zero and one should be returned
  
  assignment_vector <- rep(c('Control', 'Treatment'), each = units_per_group)
  sample(assignment_vector)
} 
```

```{r}
table(
  'Sex' = d$group, 
  'Assignment' = randomize(20)
)

table(
  'Sex' = d$group, 
  'Assignment' = randomize(20)
)

table(
  'Sex' = d$group, 
  'Assignment' = randomize(20)
)
```

```{r} 

block_ra <- function(size) { 
  ## helper function (not strictly necessary) just to 
  ## clean up the code, will be called in `randomize_block` 
  urn <- rep(c('Control', 'Treatment'), times = size/2)
  sample(urn)
}

randomize_block <- function(data_table, block_variable) { 
  ## inputs: 
  ##  - data_table, what is your data? 
  ##  - block_variable: what variable identifies your blocks? 
  data_table[ , treatment := block_ra(size = .N), by = get(block_variable)]
  
  return(data_table)
  }
```

```{r run randomize block}
dt <- randomize_block(data_table = d, block_variable = 'group')

dt[ , table(treatment, group)]
```

# Men and Women Group Variable (contd)
>randomize.blocked <- function(){
+ c(
+ sample(c(rep(0,10),rep(1,10))), #group A + sample(c(rep(0,10),rep(1,10))) #group B +)
+}
>#now groups are always balanced
>table(group, randomize.blocked())
group 0 1
Man 10 10
Woman 10 10
>table(group, randomize.blocked())
group 0 1
Man 10 10
Woman 10 10
# Simulate the Normal Study
>sim.normal.study <- function(){
+ po.control <- c(seq(from = 1, to = 20), seq(from = 51, to = 70)) + po.treatment <- po.control
+ treatment <- randomize()
+ outcomes <- po.treatment * treatment + po.control * (1–treatment) + ate <- est.ate(outcomes, treatment)
+ n.women.treatment <- table(group, treatment)[2,2]
+ return(list(ate = ate, n.women.treatment = n.women.treatment)) +}
>results <- t(replicate(1000, sim.normal.study()))
>plot(results)
 https://learn.datascience.berkeley.edu/oyster/player/latest/index.html?segmentUuid=7c4d54ad-a6b1-46bb-a1cb-e24a9815bf3a&print=true Page 2 of 3

Blocking 5/19/21, 4:52 PM
 https://learn.datascience.berkeley.edu/oyster/player/latest/index.html?segmentUuid=7c4d54ad-a6b1-46bb-a1cb-e24a9815bf3a&print=true Page 3 of 3
